Spatial and temporal variability in economic availability of residual biomass from timber harvest in west-central Alberta, Canada (1990--2015)

### 2022-11-18 analysis/ms_november/hrs_2022-11-18.Rmd
2022-11-18

abstract: |
  We use historical forest harvest data from Alberta, Canada to develop
  a description of the spatial and temporal distributions of the
  availability of forest harvest residues for the period 1990---2015.
  Using a GIS coverage of the road network in Alberta, we developed
  estimates of round-trip haul costs from each harvest location to each of
  three potential bioenergy plant locations. This information was used to
  create supply curves (or cost-availability curves) for each of the three
  potential locations and each of the 26 years. This information is used
  to develop descriptions of the variability in the delivered costs of
  forest residue for bioenergy use. These descriptions of variability are
  historical, but could be useful for potential investors in the bioenergy sector.
keywords:
  - residue
  - road network
  - supply curve
journal: "Biomass & Bioenergy"
date: "`r Sys.Date()`"
classoption: preprint, 3p, authoryear
bibliography: historicalResidueSupply.bib
linenumbers: false
numbersections: true
# Use a CSL with `citation_package = "default"`
csl: https://www.zotero.org/styles/elsevier-vancouver
output: 
  rticles::elsevier_article:
    keep_tex: true
    citation_package: default
---

@@

Liquid biofuels (*e.g.,* ethanol and biodiesel) have been produced in
commercial quantities from food crops starting in the 1980s. In some
jurisdictions policies to promote the use of these first-generation
biofuels are being phased out because of concerns that they compete with
food, and that their production may lead to land use change @Campbell2016.

Second generation biofuels rely on processes that convert
lignocellulosic materials (such a corn stover and wood) to liquid
biofuels. An advantage of this process is that the feedstock can be
waste residue left over from food or forest products production. Second
generation biofuel production is still largely in a pilot or
demonstration phase. An interesting example that will use forest
residues is the Silva Green Fuel renewable crude oil plant in Tofte,
Norway which was expected to begin operations in 2021 @Lane2018. As of the
date of writing (2022-11-07), the plant is not yet operational.

@@

Forest residues are generated as by-products of conventional forest
harvesting operations and include the treetops, branches, and
non-merchantable logs. In Canada, these residues are typically left
behind in slash piles, which are burned in winter to mitigate against
the risk of wildfires. Not all forest residues are available for
extraction, however, and estimates of residue availability should
consider ecological sustainability and technical accessibility. At a
national scale, a previous study by Yemshanov *et al.* @Yemshanov2018 estimated the
annual supply of harvestable residual biomass in Canada to be between
16.5 and 20.0 Tg/a in scenarios that included both ecological and
technical accessibility limitations. In this paper we express biomass
using SI units, and all quantities of biomass will be expressed on an
oven-dry basis: 1 Mg is equivalent to 1 oven-dry metric ton.

Despite Canada's huge potential supply of bioenergy feedstock, there are
key questions about the financial viability of forest-residue based
bioenergy investments: How much volume is spread over what area? Which
areas are best for locating a potential bioenergy plant when considering
forest residue collection and transport costs? Would forest residue
volumes be consistently available on a year-to-year basis? We attempt to
answer these questions in this paper.

There are several studies regarding plant location decisions for forest
residue-based bioenergy plants, as reviewed by Johnson *et al*. @Johnson2012.
Although these studies have improved our understanding of feedstock
availability, they tend to use limited datasets and often focus on
average levels of residue availability. A key area that has received
little attention in the bioenergy plant location literature is
consideration of feedstock supply variability over longer periods of
time. Instead of just focusing on average residue levels, a bioenergy
investor is likely to be concerned about the financial impact of
variable annual residue availability , which is the focus of our study.

@@

Biomass feedstock accounts for 40 to 60% of a bioenergy plant's total
costs (from (@Caputo2005; @Leistritz2007)  who are
cited in @Stephen2010). A number of studies have shown that collection and
transportation costs may be the biggest impediment to widespread use of
forest residues in bioenergy systems (*e.g.* @Aulakh2008). Since a forest-residue based
bioenergy plant is dependent on stemwood harvesting for feedstock, any
variability in harvesting directly influences the volume of residues
that would be available. Therefore, considering the long-term
variability associated with feedstocks in different potential locations
would be important in choosing the optimal site for locating a bioenergy
plant.

Although few studies consider long-term forest residue variability for
the purposes of optimizing the location of a potential bioenergy plant,
there are studies that consider variability for other purposes. Some
studies have considered variable feedstock supplies when investigating
whether and when to expand the production of bioenergy from burning
residues @Bolkesjo2006 @Buongiorno2011 @Galik2009. Other studies have explored the context
of variable supplies when making decisions about capital investments in
forest residue-based biorefineries @Cambero2015 @Chen2012 @Papapostolou2011. Variable
feedstock supplies are also shown to affect capital investment decisions
related to investing in a residue preprocessing facility  @Chen2012, types of storage infrastructure for bioenergy plant
feedstocks @Papapostolou2011, as well as whether to invest in biofuels versus
bioenergy, and when to invest in processing facilities @Cambero2015.

Our study contributes to the literature by considering variability in
forest residue feedstock supplies for the purpose of selecting the best
location for a bioenergy plant from alternative potential locations.
More specifically, we use precise information on the location and year
of historical timber harvest in the province of Alberta, Canada to
quantify the spatial and temporal variability of timber harvest, and
thereby relate the harvest information to spatial and temporal
variability of forest residue availability. We develop transportation
cost estimates to all of the historical timber harvest areas based on
cycle time estimates developed from a digital road network for the
province. The spatial and temporal variability of forest residues
results in variability in the transport cycle time from the residue
collection areas to the bioenergy plant, which in turn leads to
variability in delivered residue cost. From this information we develop
historical cost-availability curves for collecting and delivering
forest residues to each of the three potential bioenergy plant locations
for each of 26 years in our study period (1990--2015). When considering
locations for the potential bioenergy plants, we conduct sensitivity
analysis on 3 different bioenergy plant capacities, as defined by annual
forest residue feedstock requirements: 200 Gg/a corresponding to a small
wood pellet mill, 400 Gg/a corresponding to a large pellet mill or a
small biorefinery, and 800 Gg/a corresponding to a large biorefinery.

The methods used for this study are similar to that used by
FPInnovations for their biomass availability estimates for timber supply
areas in British Columbia (see Ref. @Blackburn2020 for an example). Like us,
FPInnovations develops cost-availability curves based on a road network
and historical harvests. However, their consideration of temporal
variability is limited to comparing two 5-year periods. As well,
FPInnovations limit their analyses to the boundaries of timber supply
areas. The only boundary considered our analysis is the provincial
boundary of Alberta.

Our analysis also shares similarities with work published by Yemshanov
*et al.* @Yemshanov2014. They develop cost-availability curves at national and
regional levels based on a national forest inventory and the location of
cogeneration facilities associated with forest products manufacturing
facilities. However, their analysis is based the reported annual volumes
of processed wood at each manufacturing plant in a single year (2010),
and the corresponding residue delivery cost.

@@

The boundaries of Canada's terrestrial ecozones can be found on
Agriculture and Agri-Food Canada's website @Canada2017. Ecozone is an input
into the NFIS individual tree biomass calculator@NFIS2021b.

@@

Most of the effort expended on this study went into aggregating
township-level summaries of data for use by the model. Much of this data
processing was the manipulation of spatial data, which was done using
the QGIS geographic information system (GIS) software @QGISorg2021 and
add-ons. Data summaries and statistical analysis were done using R
@R2021 with Rstudio @RStudioTeam2021, and tidyverse @Wickham2019.

@@

Figure 2 illustrates the road network in Alberta and estimates of cycle
time based on the average of loaded and unloaded speeds from Table 1.
The road network in the White Area of the province is very dense and
largely arranged as a grid defined by the Alberta Township Survey
System. Most of the roads in the White Area are public. The road network
in the Green Area is less dense and many of the roads are built and
maintained by resource industries (*e.g.,* forestry, petroleum, and
natural gas). The pattern of the road network is not as regular as in
the White Area.

@@

One-way haul times based on the harmonic mean travel speed were
generated using the QGIS Network Analysis Toolbox 3 plug-in (QNEAT3)
@Raffler2018, specifically its iso-area as interpolation (from point)
algorithm. In our case, we used the algorithm to determine the minimum
one-way travel time (using the harmonic mean of loaded and unloaded
speeds) along the road network from each of our centers to each cell of
a provincial level raster at a 200 m resolution. Off-road travel was
assumed to be at 5 km·h^-1^. The Dijkstra algorithm @Dijkstra1959 is the basis
for travel time minimization in QNEAT3. Cycle time is double the one-way
haul time with an extra hour added for loading and unloading.

@@

Timber Tracks produces a publication describing forestry equipment
hourly rates @TT2020. The rate reported for a tandem tractor with a
tandem trailer was 157.31 \$·hr^-1^. We assume that the tractor-trailer
combination used to haul biomass would cost a similar amount.

Therefore, the cost of transporting residual biomass was set to 13.80
\$·Mg^-1^·hr^-1^.

@@

The areas harvested by township in 2009 according to Refs. @NFIS2021a @Hermosilla2016
are shown in relationship to the three potential bioenergy plant
locations and the Green Area in Figure 1. Similar aggregated data were
created for each of the years 1990--2015. For each township, and each of
the years between 1990 and 2015, we store the area harvested (ha) for
use in our analysis. We can see from this map that a great deal of
timber harvest took place near our potential bioenergy plant locations
in 2009.

@@

The model uses a greedy algorithm whereby the lowest cost residue for a
location is extracted first. The model is summarized using pseudocode in
Figure 4.

` `  
` `  
[ Figure 4 about here ]
` `  
` `

@@

Our results assume that only one of the three potential bioenergy plants
would be built. If more than one plant was to be built, the residue
collection areas would overlap, thereby reducing the available residue
for the second plant. We will consider simultaneous location of more
than one plant in a subsequent study. Figures 5 and 6 display marginal
cycle time (h) and marginal costs (\$·Mg^-1^) for each of the 26 years
in the study period, and for the 3 potential plant locations. Curves for
1990, 2009, and 2005 are highlighted represented the years of minimum,
median, and maximum provincial harvest volume in the period. The
inter-year variation in marginal costs increases as the cumulative
amount of residue collected increases. We can also see that the relative
position of the years in terms of cost varies between the plant
locations: *e.g.,* the year 2005 was the year of highest provincial
harvest, but from a residue availability standpoint was not as good for
Hinton as it was for Edson and Whitecourt.

@@

` `  
` `  
[ Figure 6 about here]
` `  
` `

Figure 7 is another way of presenting the inter-annual variation in the
distribution of costs between the plant locations and assumed capacity:
the boxplots display the quartiles, and the minimum and maximum costs
require to supply a bioenergy plant of a given capacity at each of the
locations in each of the years. Here again, we can see the variability
increasing as plant capacity increases.

@@

Figure 8 shows the average cost (\$·Mg^-1^) for each potential location,
plant capacity, and year. Overall, the average costs in Edson and
Whitecourt are quite similar. Hinton is usually more expensive across
years and capacities. The differences become clearer at a capacity of
800 Gg·yr^-1^, for which Whitecourt is the lowest cost location for most
years.

` `  
` `  
[Figure 8 about here ]
` `  
` `

@@

Figure 9 shows the areas that would have been needed to supply a
bioenergy plant of varying capacities in Whitecourt in the year of
median harvest (2009) in relation to the proportional township area
harvested. Note that there is some residue being collected from the
White Area, but most is being collected from the Green Area.

@@

Figure 10 compares the supply areas for 2005, 2009, and 1990 (high,
medium, and low harvest years, respectively) for Whitecourt. The area
required to supply the bioenergy plant increases greatly in low timber
harvest years, particularly for the 800 Gg·yr^-1^ capacity. Similar maps
have been produced for the other locations and all 26 years and are
available at the University of Alberta's Education and Research Archive
(ERA). **will cite when location is finalized** Notice that the supply areas include areas outside the
Green Area with very little harvest. The contours indicate the cycle
time associated with each of the capacities. Because of the cost of
moving equipment such as the grinder and loader, it is unlikely that the
small, dispersed harvested areas would be accessed, but the volumes are
small enough that we ignore them. We have created animations of these
contour maps for all 26 years in the study period for the 3 locations
(available at ERA). These provide another way of visualizing the
inter-annual variation.

` `  
` `  
[ Figure 10 about here ]
` `  
` `

Whitecourt is lowest cost location for most years at all capacities.
There is a noticeable spike in costs 1999, which may be related to
salvage harvest following the 1998 Virginia Hills wildfire @Prepas2001. Note
that this spike also affects Hinton and Edson at 800 Gg·yr^-1^ capacity,
because they reach into the Whitecourt area. The average cost of
delivered forest residues are lower for the lower capacity plants, as
they do not need to collect residues from as far to meet their biomass
input requirements. In other words, the average cycle time is lower for
the lower capacity bioenergy plants, which results in a lower average
delivered residue cost.

@@

Our model demonstrates the spatial and temporal variation in the
availability of forest residue for bioenergy production. We did this
using the location of timber harvest areas in the Canadian province of
Alberta for the years 1990--2015. We conducted this study to answer
three questions critical to bioenergy investors and policy makers. We
now discuss our findings for each of these questions in turn.

**Which areas are best for locating a potential bioenergy plant when
considering residue collection and transport costs?** We focus on the
area surrounding the towns of Hinton, Edson, and Whitecourt in
west-central Alberta. These three potential bioenergy plant locations
were chosen as they host several forest products companies and are
located near forest areas with substantial harvesting activity. For each
location, we conduct sensitivity analysis on three alternative bioenergy
plant capacities as measured by residue feedstock requirements: 200
Gg·yr^-1^, 400 Gg·yr^-1^, and 800 Gg·yr^-1^. For the 200 Gg·yr^-1^
capacity the three locations are almost indistinguishable in terms of
average delivered residue cost (\~ 87 \$·Mg^-1^). At 400 Gg·yr^-1^,
Edson is the low-cost location (98.6 \$·Mg^-1^). At 800 Gg·yr^-1^,
Whitecourt is clearly the low-cost location (122 \$·Mg^-1^), however, it
is important to note that the average cost of supplying an 800 Gg·yr^-1^
plant is over 40% greater than supplying a 200 Gg·yr^-1^ plant.

**How much residue volume is spread over what area?** The residue
supply areas for the Whitecourt location shown in Figure 7 suggest that
if a 400 or 800 Gg·yr^-1^ bioenergy plant was to be constructed there
would only be enough available residue to construct one plant in our
study area. Whereas if a 200 Gg·yr^-1^ plant was deemed more viable, it
is possible that two plants could be constructed in the study area --
one in Whitecourt and one in Hinton. However, we will leave this
question to a subsequent study.

This makes for an interesting investment choice, since a single 800
Gg·yr^-1^ bioenergy plant would likely face an average delivered residue
cost that is over 40% higher than the cost for the 200 Gg·yr^-1^ plants.
Recall from above that 200 Gg·yr^-1^ is intended to represent a pellet
plant and 800 Gg·yr^-1^ is more representative of a biorefinery. Given
that dried and compacted pellets are cheaper to transport than moist and
bulky residues, it could be more viable to first transport residues to
pellet plants, and then transport the pellets from multiple pellet
plants to a biorefinery. The optimal result of this analysis will depend
on the cost of pellet production relative to the cost savings
(transportation and production) to the biorefinery from using pellets
instead of residues. Although outside the scope of this work, we plan to
conduct this analysis in the future.

**Would residue volumes be consistently available on a year-to-year
basis?** Given that a forest residue-based bioenergy plant is dependent
on forest harvesting, transportation costs increase (trucks must travel
further to meet the feedstock capacity requirement) in years when
harvesting levels are low or harvest areas are further from the
bioenergy plant. Our results showed considerable variability in both
marginal and average residue extraction costs from year-to-year for the
400 and 800 Gg·yr^-1^ capacities, suggesting that residue cost
variability is an important source of risk for medium and large
bioenergy plants. Our finding that the cost variability is lower for the
200 Gg·yr^-1^ might once again suggest that a 200 Gg·yr^-1^ is a more
appropriate scale for a forest residue-based bioenergy plant in Alberta.

@@

Although we are confident that our conclusions will hold in the future
over the long-term, it is nonetheless possible that over the long
operating life of a forest residue-based bioenergy plant, there could be
significant events that could impact residue availability over the
short-term -- *e.g.*, bad weather which prevents forest harvest and/or
residue extraction, temporary mill shutdowns, and forest fires. This
possibility makes it important for a forest residue-based bioenergy
plant to have a comprehensive feedstock risk management program. If the
bioenergy conversion technology allows it, this risk management program
could consider creating a portfolio of different sources of feedstock.
Such a portfolio approach might consider locating the bioenergy plant
where it could also access other types of feedstock -- such as
agricultural residues and/or purpose-grown feedstocks -- during periods
when availability of forest residues is low. This approach is sometimes
called "derisking" in the feedstock supply literature, whereby a supply
of "buffer" feedstock is contracted to keep the bioenergy plant running
during years when normal residue supplies are low @Golecha2016. Ref. @Zheng2021
discusses the amount and variability of production of agricultural
residue in Alberta; and Ref.\ @Shooshtarian2018 discusses the
financial viability of establishing fast-growing hybrid poplar
plantations in Alberta for use as a purpose-grown bioenergy feedstock.

When considering portfolios of different bioenergy feedstocks, an
important factor will be the average delivered cost for each feedstock.
Our estimates for average delivered forest residue costs range from
approximately 87 \$·Mg^-1^ for the 200 Gg·yr^-1^ capacity to
approximately 122 \$·Mg^-1^ for the 800 Gg·yr^-1^ capacity. Previous
studies have estimated the cost of agricultural residue (straw) to fall
within this range, with a plant gate cost in Alberta of 95.33 \$·Mg^-1^
for a plant capacity of 150 Gg·yr^-1^ @Sultana2010. Given how
close the cost of agricultural residues is to the cost of forest
residues, these two feedstocks could be a good fit for a potential
feedstock portfolio, assuming the bioenergy plant could use both types
of feedstock. However, a previous study estimated that purpose-grown
hybrid poplar in Alberta has a much higher delivered cost of 202
\$·Mg^-1^ (125 \$·Mg^-1^ to grow the trees, plus 77 \$·Mg^-1^ to
harvest, grind, and transport the trees to the bioenergy plant)
@Shooshtarian2018, suggesting that hybrid poplar would have a
more limited role (if used at all) within a feedstock portfolio. In the
future we hope to build upon this research by optimizing the allocation
of different feedstocks within a bioenergy feedstock portfolio.

@@

We work with a group of talented people. Thanks to Marty Luckert, Feng
Qiu, Grant Hauer, and Wenbei Zhang for helpful discussion.

>>>>

abstract: |
  We use historical forest harvest data from Alberta, Canada to develop
  a description of the spatial and temporal distributions of the
  availability of forest harvest residues for the period 1990---2015.
  Using a GIS coverage of the road network in Alberta, we developed
  estimates of round-trip haul costs from each harvest location to each of
  three potential bioenergy plant locations. This information was used to
  create cost curves for each of the three
  potential locations and each of the 26 years. This information is used
  to develop descriptions of the variability in the delivered costs of
  forest residue for bioenergy use. These descriptions of variability are
  historical, but could be useful for potential investors in the bioenergy sector.
keywords:
  - residue
  - road network
  - supply curve
journal: "Biomass & Bioenergy"
date: "`r Sys.Date()`"
classoption: preprint, 3p, authoryear
bibliography: historicalResidueSupply.bib
linenumbers: false
numbersections: true
# Use a CSL with `citation_package = "default"`
csl: https://www.zotero.org/styles/elsevier-vancouver
output: 
  rticles::elsevier_article:
    keep_tex: true
    keep_md: true
    citation_package: default
---

@@

Liquid biofuels (*e.g.,* ethanol and biodiesel) have been produced in
commercial quantities from food crops since the 1980s. In some
jurisdictions, policies to promote the use of these first-generation
biofuels are being phased out because of concerns that they compete with
food, and that their production may lead to land use change @Campbell2016.

Second generation biofuels rely on processes that convert
lignocellulosic materials (such a corn stover and wood) to liquid
biofuels. An advantage of this process is that the feedstock can be
waste residue from food or forest products production. Second
generation biofuel production is still largely in a pilot or
demonstration phase. An interesting example that will use forest
residues is the Silva Green Fuel renewable crude oil plant in Tofte,
Norway which was expected to begin operations in 2021 @Lane2018. As of the
date of writing (2022-11-07), the plant is not yet operational.

@@

Forest residues are generated as by-products of conventional forest
harvesting operations and include the treetops, branches, and
non-merchantable logs. In Canada, these residues are typically left
behind in slash piles, which are burned in winter to mitigate against
the risk of wildfires. Not all forest residues are available for
extraction, however, and estimates of residue availability should
consider ecological sustainability and technical accessibility. At a
national scale, a previous study by Yemshanov *et al.* @Yemshanov2018 estimated the
annual supply of harvestable residual biomass in Canada to be between
16.5 and 20.0 Tg/a in scenarios that included both ecological and
technical accessibility limitations. ^[In this paper we express biomass
using SI units, and all quantities of biomass will be expressed on an
oven-dry basis: 1 Mg is equivalent to 1 oven-dry metric ton.]

Despite Canada's huge potential supply of bioenergy feedstock, there are
key questions about the financial viability of forest-residue based
bioenergy investments: How much volume is spread over what area? Which
areas are best for locating a potential bioenergy plant when considering
forest residue collection and transport costs? 
Will forest residue
volumes be consistently available on a year-to-year basis? We attempt to
answer these questions in this paper.

There are several studies regarding plant location decisions for forest
residue-based bioenergy plants, as reviewed by Johnson *et al*. @Johnson2012.
Although these studies have improved our understanding of feedstock
availability, they tend to use limited datasets and often focus on
average levels of residue availability. A key area that has received
little attention in the bioenergy plant location literature is
consideration of feedstock supply variability over longer periods of
time. Instead of just focusing on average residue levels, a bioenergy
investor is likely to be concerned about the financial impact of
variable annual residue availability, which is the focus of our study.

@@

Biomass feedstock accounts for 40 to 60% of a bioenergy plant's total
costs (from (@Caputo2005; @Leistritz2007)  who are
cited in @Stephen2010). A number of studies have shown that collection and
transportation costs may be the biggest impediment to widespread use of
forest residues in bioenergy systems (*e.g.* @Aulakh2008). Since a forest-residue based
bioenergy plant is dependent on stemwood harvesting for feedstock, any
variability in harvesting directly influences the volume of residues
that would be available. Therefore, considering the long-term
variability associated with feedstock in different potential locations
would affect the optimal siting for a bioenergy
plant.

Although few studies consider long-term forest residue variability for
the purposes of optimizing the location of a potential bioenergy plant,
there are studies that consider variability for other purposes. Some
studies have considered variable feedstock supplies when investigating
whether and when to expand the production of bioenergy from burning
residues @Bolkesjo2006 @Buongiorno2011 @Galik2009. Other studies have explored the context
of variable supplies when making decisions about capital investments in
forest residue-based biorefineries @Cambero2015 @Chen2012 @Papapostolou2011. Variable
feedstock supplies are also shown to affect capital investment decisions
related to investing in a residue preprocessing facility  @Chen2012, types of storage infrastructure for bioenergy plant
feedstock @Papapostolou2011, as well as whether to invest in biofuels versus
bioenergy, and when to invest in processing facilities @Cambero2015.

Our study contributes to the literature by considering variability in
forest residue feedstock supplies for the purpose of selecting the best
location for a bioenergy plant from alternative potential locations.
More specifically, we use precise information on the location and year
of historical timber harvest in the province of Alberta, Canada to
quantify the spatial and temporal variability of timber harvest, and
thereby relate the harvest information to spatial and temporal
variability of forest residue availability. We develop transportation
cost estimates to all of the historical timber harvest areas based on
cycle time estimates developed from the road network for the
province. The spatial and temporal variability of forest residues
results in variability in the transport cycle time from the residue
collection areas to the bioenergy plant, which in turn leads to
variability in delivered residue cost. From this information we develop
historical cost curves for collecting and delivering
forest residues to each of the three potential bioenergy plant locations
for each of 26 years in our study period (1990--2015). When considering
locations for the potential bioenergy plants, we conduct sensitivity
analysis on 3 different bioenergy plant capacities, as defined by annual
forest residue feedstock requirements: 200 Gg/a corresponding to a small
wood pellet mill, 400 Gg/a corresponding to a large pellet mill or a
small biorefinery, and 800 Gg/a corresponding to a large biorefinery.

The methods used for this study are similar to that used by
FPInnovations for their biomass availability estimates for timber supply
areas in British Columbia (see Ref. @Blackburn2020 for an example). Like us,
FPInnovations develops cost curves based on a road network
and historical harvests. However, their consideration of temporal
variability is limited to comparing two 5-year periods. As well,
FPInnovations limit their analyses to the boundaries of timber supply
areas. The only boundary considered our analysis is the provincial
boundary of Alberta.

Our analysis also shares similarities with work published by Yemshanov
*et al.* @Yemshanov2014. They develop cost curves at national and
regional levels based on a national forest inventory and the location of
cogeneration facilities associated with forest products manufacturing
facilities. However, their analysis is based the reported annual volumes
of processed wood at each manufacturing plant in a single year (2010),
and the corresponding residue delivery cost.

@@

The boundaries of Canada's terrestrial ecozones can be found on
Agriculture and Agri-Food Canada's website @Canada2017. Ecozone is an input
into the NFIS individual tree biomass calculator @NFIS2021b.

@@

Most of the effort expended on this study went into aggregating
township-level summaries of data for use by the model. Much of this data
processing was the manipulation of spatial data, which was done using
the QGIS geographic information system (GIS) software @QGISorg2021 and
add-ons. Data summaries and statistical analysis were done using R
@R2021 with RStudio @RStudioTeam2021, and tidyverse @Wickham2019.

@@

Figure 2 illustrates the road network in Alberta and estimates of cycle
time based on the average of loaded and unloaded speeds from Table 1.
The road network in the White Area of the province is very dense and
largely arranged as a grid defined by the Alberta Township Survey
System. Most of the roads in the White Area are public. The road network
in the Green Area is less dense and many of the roads are built and
maintained by resource industries (*e.g.,* forestry, petroleum, and
natural gas). The pattern of the road network is less regular than in
the White Area.

@@

One-way haul times based on the harmonic mean travel speed were
generated using the QGIS Network Analysis Toolbox 3 plug-in (QNEAT3)
@Raffler2018, specifically its iso-area as interpolation (from point)
algorithm. In our case, we used the algorithm to determine the minimum
one-way travel time (using the harmonic mean of loaded and unloaded
speeds) along the road network from each of our centers to each cell of
a provincial level raster at a 200 m resolution. Off-road travel was
assumed to be at 5 km/h. The Dijkstra algorithm @Dijkstra1959 is the basis
for travel time minimization in QNEAT3. Cycle time is double the one-way
haul time with an extra hour added for loading and unloading.

@@

Timber Tracks produces a publication describing forestry equipment
hourly rates @TT2020. The rate reported for a tandem tractor with a
tandem trailer was 157.31 \$/h. We assume that the tractor-trailer
combination used to haul biomass would cost a similar amount.

Therefore, the cost of transporting residual biomass was set to 13.80
(\$·Mg^-1)/h.

@@

The areas harvested by township in 2009 according to Refs. @NFIS2021a @Hermosilla2016
are shown in relationship to the three potential bioenergy plant
locations in Figure 1. Similar aggregated data were
created for each of the years 1990--2015. For each township, and each of
the years between 1990 and 2015, we store the area harvested (ha) for
use in our analysis. We can see from this map that a great deal of
timber harvest took place near our potential bioenergy plant locations
in 2009.

@@

The model uses a greedy algorithm whereby the lowest cost residue for a
location is extracted first. The model is summarized using pseudocode in
Figure 3.

` `  
` `  
[ Figure 3 about here ]
` `  
` `

@@

Our analysis assumes that only one of the three potential bioenergy plants
would be built. If more than one plant was to be built, the residue
collection areas could overlap, thereby reducing the available residue
for the second plant. We will consider simultaneous location of more
than one plant in a subsequent study. Figures 4 and 5 display marginal
cycle time (h) and marginal costs (\$·Mg^-1^) for each of the 26 years
in the study period, and for the 3 potential plant locations. Curves for
1990, 2009, and 2005 are highlighted represented the years of minimum,
median, and maximum provincial harvest volume in the period. The
inter-year variation in marginal costs increases as the cumulative
amount of residue collected increases. We can also see that the relative
position of the years in terms of cost varies between the plant
locations: *e.g.,* the year 2005 was the year of highest provincial
harvest, but from a residue availability standpoint was not as good for
Hinton as it was for Edson and Whitecourt.

` `  
` `  
[ Figure 4 about here]
` `  
` `

@@

Figure 6 is presents the inter-annual variation in the
distribution of costs between the plant locations and assumed capacity in another way.
The boxplots display the quartiles, and the minimum and maximum costs
required to supply a bioenergy plant of a given capacity at each of the
locations in each of the years. Here again, we can see the variability
increasing as plant capacity increases.

` `  
` `  
[Figure 6 about here ]
` `  
` `

Figure 7 shows the average cost (\$·Mg^-1^) for each potential location,
plant capacity, and year. Overall, the average costs in Edson and
Whitecourt are quite similar. Hinton is usually more expensive across
years and capacities. The differences become clearer at a capacity of
800 Gg/a, for which Whitecourt is the lowest cost location for most
years.

@@

@@

Figure 8 shows the areas that would have been needed to supply a
bioenergy plant of varying capacities in Whitecourt in the year of
median harvest (2009) in relation to the proportional township area
harvested. Note that there is some residue being collected from the
White Area, but most is being collected from the Green Area.

` `  
` `  
[ Figure 8 about here ]
` `  
` `

Figure 9 compares the supply areas for 2005, 2009, and 1990 (high,
medium, and low harvest years, respectively) for Whitecourt. The area
required to supply the bioenergy plant increases greatly in low timber
harvest years, particularly for the 800 Gg/a capacity. Similar maps
have been produced for the other locations and all 26 years and are
available at the University of Alberta's Education and Research Archive
(\url{https://doi.org/10.7939/r3-1257-j192}). Notice that the supply areas include areas outside the
Green Area with very little harvest. The contours indicate the cycle
time associated with each of the capacities. Because of the cost of
moving equipment such as the grinder and loader, it is unlikely that the
small, dispersed harvested areas would be accessed, but the volumes are
small enough that we ignore them. We have created animations of these
contour maps for all 26 years in the study period for the 3 locations
(available at ERA). These provide another way of visualizing the
inter-annual variation.

@@

Whitecourt is lowest cost location for most years at all capacities.
There is a noticeable spike in costs 1999, which may be related to
salvage harvest following the 1998 Virginia Hills wildfire @Prepas2001. Note
that this spike also affects Hinton and Edson at 800 Gg/a capacity,
because they reach into the Whitecourt area. The average cost of
delivered forest residues are lower for the lower capacity plants, as
they do not need to collect residues from as far to meet their biomass
input requirements. In other words, the average cycle time is lower for
the lower capacity bioenergy plants, which results in a lower average
delivered residue cost.

@@

Our results demonstrate the spatial and temporal variation in the
availability of forest residue for bioenergy production. This was done
using the location of timber harvest areas in
Alberta, Canada for the years 1990--2015. We conducted this study to answer
three questions critical to bioenergy investors and policy makers. We
now discuss our findings for each of these questions in turn.

**Which areas are best for locating a potential bioenergy plant when
considering residue collection and transport costs?** We focus on the
area surrounding the towns of Hinton, Edson, and Whitecourt in
west-central Alberta. These three potential bioenergy plant locations
were chosen as they host several forest products companies and are
located near forest areas with substantial harvesting activity. For each
location, we conduct sensitivity analysis on three alternative bioenergy
plant capacities as measured by residue feedstock requirements: 200
Gg/a, 400 Gg/a, and 800 Gg/a. For the 200 Gg/a
capacity the three locations are almost indistinguishable in terms of
average delivered residue cost (\~ 87 \$·Mg^-1^). At 400 Gg/a,
Edson is the low-cost location (98.6 \$·Mg^-1^). At 800 Gg/a,
Whitecourt is clearly the low-cost location (122 \$·Mg^-1^), however, it
is important to note that the average cost of supplying an 800 Gg/a
plant is over 40% greater than supplying a 200 Gg/a plant.

**How much residue volume is spread over what area?** The residue
supply areas for the Whitecourt location shown in Figure 7 suggest that
if a 400 or 800 Gg/a bioenergy plant was to be constructed there
would only be enough available residue to construct one plant in our
study area. Whereas if a 200 Gg/a plant was deemed more viable, it
is possible that two plants could be constructed in the study area --
one in Whitecourt and one in Hinton. However, we will leave this
question to a subsequent study.

This makes for an interesting investment choice, since a single 800
Gg/a bioenergy plant would likely face an average delivered residue
cost that is over 40% higher than the cost for the 200 Gg/a plants.
Recall from above that 200 Gg/a is intended to represent a pellet
plant and 800 Gg/a is more representative of a biorefinery. Given
that dried and compacted pellets are cheaper to transport than moist and
bulky residues, it could be more viable to first transport residues to
pellet plants, and then transport the pellets from multiple pellet
plants to a biorefinery. The optimal result of this analysis will depend
on the cost of pellet production relative to the cost savings
(transportation and production) to the biorefinery from using pellets
instead of residues. Although outside the scope of this work, we plan to
conduct this analysis in the future.

**Will residue volumes be consistently available on a year-to-year
basis?** Given that a forest residue-based bioenergy plant is dependent
on forest harvesting, transportation costs increase (trucks must travel
further to meet the feedstock capacity requirement) in years when
harvesting levels are low or harvest areas are further from the
bioenergy plant. Our results showed considerable variability in both
marginal and average residue extraction costs from year-to-year for the
400 and 800 Gg/a capacities, suggesting that residue cost
variability is an important source of risk for medium and large
bioenergy plants. Our finding that the cost variability is lower for the
200 Gg/a might once again suggest that a 200 Gg/a is a more
appropriate scale for a forest residue-based bioenergy plant in Alberta.

@@

Although we are confident that our conclusions will hold in the future
over the long-term, it is nonetheless possible that over the long
operating life of a forest residue-based bioenergy plant, there could be
significant events that could impact residue availability over the
short-term -- *e.g.*, bad weather which prevents forest harvest and/or
residue extraction, temporary mill shutdowns, and forest fires. This
possibility makes it important for a forest residue-based bioenergy
plant to have a comprehensive feedstock risk management program. If the
bioenergy conversion technology allows it, this risk management program
could consider creating a portfolio of different sources of feedstock.
Such a portfolio approach might consider locating the bioenergy plant
where it could also access other types of feedstock -- such as
agricultural residues and/or purpose-grown feedstock -- during periods
when availability of forest residues is low. This approach is sometimes
called "derisking" in the feedstock supply literature, whereby a supply
of "buffer" feedstock is contracted to keep the bioenergy plant running
during years when normal residue supplies are low @Golecha2016. Ref. @Zheng2021
discusses the amount and variability of production of agricultural
residue in Alberta; and Ref.\ @Shooshtarian2018 discusses the
financial viability of establishing fast-growing hybrid poplar
plantations in Alberta for use as a purpose-grown bioenergy feedstock.

When considering portfolios of different bioenergy feedstock, an
important factor will be the average delivered cost for each feedstock.
Our estimates for average delivered forest residue costs range from
approximately 87 \$·Mg^-1^ for the 200 Gg/a capacity to
approximately 122 \$·Mg^-1^ for the 800 Gg/a capacity. Previous
studies have estimated the cost of agricultural residue (straw) to fall
within this range, with a plant gate cost in Alberta of 95.33 \$·Mg^-1^
for a plant capacity of 150 Gg/a @Sultana2010. Given how
close the cost of agricultural residues is to the cost of forest
residues, these two feedstock could be a good fit for a potential
feedstock portfolio, assuming the bioenergy plant could use both types
of feedstock. However, a previous study estimated that purpose-grown
hybrid poplar in Alberta has a much higher delivered cost of 202
\$·Mg^-1^ (125 \$·Mg^-1^ to grow the trees, plus 77 \$·Mg^-1^ to
harvest, grind, and transport the trees to the bioenergy plant)
@Shooshtarian2018, suggesting that hybrid poplar would have a
more limited role (if used at all) within a feedstock portfolio. We plan to build upon this research by optimizing the allocation
of different feedstock within a bioenergy feedstock portfolio.

@@

Thanks go to Feng
Qiu, Grant Hauer, and Wenbei Zhang for helpful discussion.
